Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife

### 2024-05-14 analysis/paper/paper.Rmd
delete old figures, reformat study area

```{=html}
<!--        TO DO: consider these comments from Kyle: Need a description about road development. In BC, forestry roads are developed by forest licensees under a timber harvesting permit. Various considerations go into road development including soils, grade, and economics. What type/classification of roads are you looking at (primitive roads, paved roads, etc). What’s the maximum grade? ---can find a lot of this verbiage in the Interior Appraisal Manual. How are they maintained – do they last forever -see Auditor General Report 
•        How much of the Revelstoke TSA overlaps with caribou range? Elevation? Climate? Flora/Fauna?
•        Maybe add a table of descriptive statistics presenting information about roads in the Revelstoke TSA and the other TSAs you mention? By Caribou Range? – road density, road length, cutblock area, cutblock density, etc
-->
```
```{r setup2}
library(sf)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(ggpubr)
library(MetBrewer)
library(here)
library(tmap)
library(cowplot)

@@

<!---BLOCK_MULTICOL_START--->

@@

canada_overlay <- tm_shape(canada, projection = "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")+
  tm_borders(col="black")+
  tm_shape(st_buffer(tsaBoundary, 10000))+
  tm_polygons(border.col = met.brewer(pal_nm, 1), col = met.brewer(pal_nm, 1))+
  tm_layout(frame=TRUE)

tsa <- tm_shape(tsaCost, name = "Cost")+
  tm_raster(style = "cont", title = "Cost",palette="Greys")+
  tm_shape(st_union(tsaBoundary))+
  tm_borders(col = "black")+
  tm_compass(position = c(0.3, 0.11), size = 1)+
  tm_scale_bar(position = c("center", "bottom"), text.size = 1)+
  tm_layout(legend.position = c("left", "bottom"),
            legend.bg.color = "white")

vp <- grid::viewport(x=0.965, y=0.945, width = 0.35, height=0.2,
               just=c("right", "top"))

tmap_save(tsa, filename = here("analysis/figures/Figure1_StudyArea.png"),
  dpi = 600, insets_tm = canada_overlay, insets_vp = vp,
  height = 6*0.9, width = fig_widths["single"]*0.9, units = "in")

tsaDEM <- tm_shape(dem, name = "Elevation")+
  tm_raster(style = "cont", title = "Elevation",palette="Greys")+
  tm_shape(st_union(tsaBoundary))+
  tm_borders(col = "black")+
  # tm_compass(position = c(0.3, 0.11), size = 1)+
  # tm_scale_bar(position = c("center", "bottom"), text.size = 1)+
  tm_layout(legend.position = c("left", "bottom"),
            legend.bg.color = "white")

vp2 <- grid::viewport(x=0.965, y=0.945, width = 0.35, height=0.2,
                      just=c("right", "top"))

both_maps <- tmap_arrange(tsa, tsaDEM, outer.margins = 0)

tsa_grob <- tmap_grob(tsa)
tsaDEM_grob <- tmap_grob(tsaDEM)
canada_overlay_grob <- tmap_grob(canada_overlay)

out_map <- plot_grid(tsa_grob, tsaDEM_grob)+
    draw_grob(canada_overlay_grob, x=0.965, y=0.945,
            width = 0.35, height=0.2,
            hjust= 1, vjust = 1)

tmap_save(both_maps, filename = here("analysis/figures/Figure1_StudyArea_both.tiff"),
  dpi = 600, insets_tm = canada_overlay, insets_vp = vp2,
  height = 6*0.9, width = fig_widths["two"], units = "in")

tmap_save(tsaDEM, filename = here("analysis/figures/Figure1_StudyAreaDEM.png"),
  dpi = 600, insets_vp = vp,
  height = 6*0.9, width = fig_widths["single"]*0.9, units = "in")

# try with ggspatial

library(ggspatial)
library(ggplot2)

ggplot() + 
  layer_spatial(dem) +
  geom_sf(data = st_union(tsaBoundary), fill = NA) +
  annotation_scale() +
  scale_fill_continuous(na.value = NA)+
  theme_void()
```

```{r study-area1, fig.cap= "see next fig"}

knitr::include_graphics(here("analysis/figures/Figure1_StudyArea.png"))

```{r study-area2}

knitr::include_graphics(here("analysis/figures/Figure1_StudyAreaDEM.png"))
```

<!---BLOCK_MULTICOL_STOP{widths: [3.6,3.6]}--->

@@

png::writePNG(array(0, dim = c(1,1,4)), here("analysis/figures/empty.png"))
knitr::include_graphics(here("analysis/figures/empty.png"))
```

@@

To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023]. The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022]. The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].

@@

tmap_save(tmap_arrange(allMaps, ncol = 3),
          here("analysis/figures/projection_methods_figure.png"),
          dpi = 300, height = 8, width = fig_widths["two"])
```

```{r proj-meth-fig, fig.cap="Comparison of road projection method variants (Table 2). The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks.The grade penalty method for setting edge weights (top row) encourages roads to follow slope contours, and yields fewer obviously unrealistic 90 degree turns in steep terrain than our simple cost method (middle row). Selecting a single landing at the centroid of each cutblock (bottom left) creates too few roads within cutblocks. In this 1 ha resolution example the Hardy QGIS plugin (bottom middle) creates more curved roads than our simple cost method (middle row), and does not follow slope contours like the grade penalty method (top row). Landing density is high for all except the centroid and Hardy variants."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
```

>>>>

```{=html}
<!--        TO DO: consider these comments from Kyle: Need a description about road development. In BC, forestry roads are developed by forest licensees under a timber harvesting permit. Various considerations go into road development including soils, grade, and economics. What type/classification of roads are you looking at (primitive roads, paved roads, etc). What’s the maximum grade? ---can find a lot of this verbiage in the Interior Appraisal Manual. How are they maintained – do they last forever -see Auditor General Report 
•        How much of the Revelstoke TSA overlaps with caribou range? Elevation? Climate? Flora/Fauna?
•        Maybe add a table of descriptive statistics presenting information about roads in the Revelstoke TSA and the other TSAs you mention? By Caribou Range? – road density, road length, cutblock area, cutblock density, etc
-->
```
```{r setup2}
library(sf)
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(purrr)
library(MetBrewer)
library(here)
library(tmap)
library(cowplot)
library(ggspatial)

@@

@@

# using ggspatial because tmap doesn't work with both multiplots and insets
tsaBound <- st_union(tsaBoundary)

dem_map <- ggplot() + 
  layer_spatial(dem) +
  geom_sf(data = tsaBound, fill = NA) +
  annotation_scale(width_hint = 0.4) +
  scale_fill_continuous(low = "white", high = "black",
                        na.value = NA)+
  theme_void()+
  theme(plot.background = element_rect(colour = NA, fill = "white"))+
  labs(fill = "Elevation")

cost_map <- ggplot() + 
  layer_spatial(tsaCost) +
  geom_sf(data = tsaBound, fill = NA) +
  annotation_north_arrow(style = north_arrow_orienteering(fill = c("black", "black")), height = unit(1, "cm"), width = unit(0.75, "cm")) +
  scale_fill_continuous(low = "white", high = "black",
                        na.value = NA)+
  theme_void()+
  theme(plot.background = element_rect(colour = NA, fill = "white"))+
  labs(fill = "Cost")

can_map <- ggplot() + 
  geom_sf(data = canada %>% st_transform(crs = "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"), fill = NA) +
  geom_sf(data = st_buffer(tsaBound, 10000) %>% st_transform(crs = "+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs"), 
          fill = met.brewer(pal_nm, 1)) +
  theme_void()+
  theme(panel.border = element_rect(colour = "black", fill = NA))

out_map <- plot_grid(cost_map, dem_map)+
  draw_plot(can_map, x=1.09, y=0.945,
            width = 0.35, height=0.2,
            hjust= 1, vjust = 1)

save_plot(here("analysis/figures/Figure1_StudyArea_both.tiff"), 
          out_map,
          dpi = knitr::opts_chunk$get("dpi"),
          base_width = fig_widths["two"], 
          base_height = 6*0.9)
```

@@

knitr::include_graphics(here("analysis/figures/Figure1_StudyArea_both.tiff"))
```

@@

To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023]. The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csárdi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022]. The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].

@@

tmap_save(tmap_arrange(allMaps, ncol = 3),
          here("analysis/figures/projection_methods_figure.tiff"),
          dpi = 300, height = 8, width = fig_widths["two"])
```

```{r proj-meth-fig, fig.cap="Comparison of road projection method variants (Table 2). The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks.The grade penalty method for setting edge weights (top row) encourages roads to follow slope contours, and yields fewer obviously unrealistic 90 degree turns in steep terrain than our simple cost method (middle row). Selecting a single landing at the centroid of each cutblock (bottom left) creates too few roads within cutblocks. In this 1 ha resolution example the Hardy QGIS plugin (bottom middle) creates more curved roads than our simple cost method (middle row), and does not follow slope contours like the grade penalty method (top row). Landing density is high for all except the centroid and Hardy variants."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.tiff"))
```
